---
title: 'Lab 01 - Functions and Vectors'
output:
  html_document:
    theme: readable
    df_print: paged
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=F, warning=F)
library( dplyr )
library( pander )
```

#### [CPP 526: Data Science I](https://ds4ps.org/cpp-526-fall-2019/)

<br>

*This lab offers practice with some basic R functions for summarizing vectors.*

I have provided you with an RMD template for this lab: [Download RMD Template](https://www.dropbox.com/s/43dj0oszgg34clc/Lab-01-LastName.Rmd?dl=1)


If you are having trouble with your RMD file, try the [RMD File Styles and Knitting Tips](https://ds4ps.org/cpp-526-fall-2019/labs/r-markdown-files.html) manual.

<br><br>

## Functions

You will use the following functions for this lab: 

```{r,eval=F}
names()                 # variable names
head()                  # preview dataset
length()                # vector length (number of elements)
dim(), nrow(), ncol()   # dataset dimensions
sum(), summary()        # summarize numeric vectors
table()                 # summarize factors / character vectors
```

<br><br>



# Data

This lab uses **city tax parcel data** from Syracuse, NY.

Each house, business, park or empty lot in the city sits on a parcel of land, and the tax rate for that parcel is determined by the city. 

All of the characteristics of each parcel is stored in a central database, along with information about the land use classifications, owners, and whether the property is up-to-date on taxes and water bills. 

Each row represents one tax parcel, and all tax parcels in the city are recorded in the dataset. 

[**DATA DICTIONARY**](https://ds4ps.github.io/Data-Science-Class/DATA/syr_parcels.html)

You can load the dataset as follows: 

```{r}
URL <- "https://raw.githubusercontent.com/DS4PS/Data-Science-Class/master/DATA/syr_parcels.csv"
dat <- read.csv( URL, stringsAsFactors=FALSE )
```

Preview the dataset using the **head()** function:

```{r, echo=T}
head( dat, 10 ) # show first ten rows of data
```



<br><br>

# Lab Instructions

Answer the following questions in your lab.  

1. How many tax parcels are in Syracuse (dataset dimensions)? 
2. How many acres of land are in syracuse (sum over a numeric vector)?
3. How many vacant BUILDINGS are there in Syracuse (sum over a logical vector)?
4. What proportion of parcels are tax exempt ?
5. Which neighborhood contains the most tax parcels (table with one variable)?
6. Which neighborhood contains the most vacant LOTS (table with two variables)? 


**Note:**

(1) Referencing variables in R requires both the **dataset name** and **variable name**, separated by the `$` operator:

```{r, eval=F}
summary( dat$acres )
```

(2) When you apply a **sum()** function to a numeric vector it returns the sum of all elements in the vector. 

```{r}
sum( c(10,20,5) )
```

When you apply a **sum()** function to a logical vector, it will count all of the TRUEs:

```{r}
x <- c(T,T,F,F,F)
sum( x )
sum( x ) / length( x ) # proportion of trues
```


(3) R wants to make sure you are aware of missing values, so it will return NA (not available) for functions performed on vectors with missing values. 

Add the 'NA remove' argument (`na.rm=TRUE`) to functions to ignore missing values: 

```{r, eval=F}
sum( dat$star, na.rm=T )
```





<br><br>

# Submission Instructions

When you have completed your assignment, knit your RMD file to generate your rendered HTML file. Platforms like BlackBoard and Canvas often disallow you from submitting HTML files when there is embedded computer code, so create a zipped folder with both the RMD and HTML files.

Login to Canvas at <http://canvas.asu.edu> and navigate to the assignments tab in the course repository. Upload your zipped folder to the appropriate lab submission link.

Remember to:

* name your files according to the convention: **Lab-##-LastName.Rmd**
* show your solution, include your code.
* do not print excessive output (like a full data set).
* follow appropriate style guidelines (spaces between arguments, etc.).

See [Google's R Style Guide](https://google.github.io/styleguide/Rguide.xml) for examples.

## Notes on Knitting

Note that when you knit a file, it starts from a blank slate. You might have packages loaded or datasets active on your local machine, so you can run code chunks fine. But when you knit you might get errors that functions cannot be located or datasets don't exist. Be sure that you have included chunks to load these in your RMD file.

Your RMD file will not knit if you have errors in your code. If you get stuck on a question, just add `eval=F` to the code chunk and it will be ignored when you knit your file. That way I can give you credit for attempting the question and provide guidance on fixing the problem.


